Robust inference of neutron-star parameters from thermonuclear burst

### 2019-08-26 doc/ms.tex
Updated observe method to use correct scalings for GR. Modified the compare method to use the observe method (instead of it's own "custom" version). Updated the plot routine to allow arguments to be passed to the plot commands, and also set up a convention for colors for observed and simulated bursts
Added extra routines designed for inferring neutron star properties from bursts, copied from Inferring\ composition.ipynb

@@

@@

% \title{Thermonuclear burst model-observation comparisons with {\sc concord}}
\title{Robust inference of neutron-star parameters from thermonuclear burst lightcurves}

@@

The fundamental observables that we measure include the recurrence time $\Delta t$; the persistent flux $F_{\rm per}$, a measure of the accretion rate; and the burst bolometric flux lightcurve, $(t_i,F_{i})$. From the burst lightcurve we may integrate to estimate the burst fluence, $E_b$, which may be used to estimate the $\alpha$-value, the ratio of the burst to persistent flux: 
\begin{equation}
\alpha = \frac{\Delta t F_{\rm per} c_{\rm bol}}{E_B}
\end{equation}
where $c_{\rm bol}$ is the bolometric correction giving the inverse fraction of total persistent flux emitted in the instrumental band.

@@

% The corrections described below
However, since the model is to be transformed by a different value of $(1+z)_{\rm fit}$, we must take care to consistently transform the model parameters in accordance with this choice.
%
We can achieve this objective as follows. The surface gravity $g$ adopted for the model, and the trial redshift $(1+z)_{\rm fit}$ imply a different neutron star mass and radius than adopted for the model: 
%
% We note that the combination of surface gravity $g$ (assumed for the model) and adopted $1+z$  uniquely specifies the neutron star mass $M_{\rm NS}$ and radius $R_{\rm NS}$:
\begin{eqnarray}
% as used in function calc_mr in burstclass.py
R_{\rm NS,fit} &=& c^2\frac{(1+z)_{\rm fit}^2-1}{2g(1+z)_{\rm fit}}\\
M_{\rm NS,fit} &=& \frac{gR_{\rm NS}^2}{G(1+z)_{\rm fit}}
\end{eqnarray}
Thus, by identifying the optimal value of $(1+z)_{\rm fit}$ for comparison to a particular observation, we can constrain the mass and radius, at a fixed $g$. [{\bf so which one should I use? and should I make any other changes to the recipe?}]

@@

Given an estimate of $Q_{\rm nuc}$, the average hydrogen fraction $\langle X\rangle$ in the fuel layer at ignition has been estimated to date using the relation $Q_{\rm nuc}=1.6 + 4\langle X\rangle$~MeV/nucleon \cite[e.g.][and references therein]{gal03d}. This expression includes $\approx35$\% losses
attributed to neutrino emission \cite[]{fuji87}. However, the 35\% value applies only to $\beta$-decays in the much more extensive {\it rp}-process burning chain, and recent {\sc kepler} simulations indicate instead an approximation of $Q{\rm nuc}=0.96+6.6\langle X\rangle$ (Goodwin et al. 2018, in preparation). [{\it comment on improved quadratic expression}]
%
Thus, by substituting in equation \ref{qnuc} we can estimate 
\begin{equation}
\langle X\rangle 
% In [42]: c.c**2/(6.6*u.MeV/c.m_p).to("m**2/s**2")
% Out[42]: <Quantity 142.16243658783114>
% = \frac{c^2z/\alpha}{6.6\ {\rm Mev/nucleon}}\frac{\xi_b}{\xi_p} - 0.145
= z\frac{142}{\alpha}\frac{\xi_b}{\xi_p} - 0.145 \label{xbar}
\end{equation}
[{\it comment on the degree of error introduced by ignoring anisotropy etc.}] 
%
Clearly this expression is only applicable for $\alpha$-values up to some limit, which we calculate as
\begin{equation}
%In [43]: c.c**2/(1.*u.MeV/c.m_p).to("m**2/s**2")/6.6/0.145
%Out[43]: <Quantity 980.4305971574562>
\alpha \leq 980\, z \frac{\xi_b}{\xi_p}
\end{equation}
Interestingly, this limit is $\approx70$\% larger than the equivalent value derived for the old expression for $Q_{\rm nuc}$. This result suggests that larger observed values of $\alpha$ may be accommodated without resorting to explanations including incomplete burning of burst fuel \cite[e.g.][]{bcatalog}.

@@

The hydrogen fraction at ignition $\langle X\rangle$ and the burst recurrence time may then be used to estimate the hydrogen fraction in the fuel, $X_0$. The H-fraction at the base of the fuel layer is reduced steadily by $\beta$-limited hot CNO burning, and will completely exhaust the accreted hydrogen in a time \cite[]{lampe16}:
\begin{equation}
t_{\rm CNO} =9.8\left(\frac{X_0}{0.7}\right) \left(\frac{Z_{\rm CNO}}{0.02}\right)^{-1}
\end{equation}
measured in the neutron star frame.
The hydrogen fraction at the base will thus be $X_0[1-\Delta t/(1+z)t_{\rm CNO}]$, 
% [{\it should this expression include time dilation?}] -- yes
so that the average H-fraction in the layer will be $\langle X \rangle = X_0[1-0.5\Delta t/(1+z)t_{\rm CNO}]$. We can combine this expression with equation \ref{xbar} to give
\begin{eqnarray}
X_0 & = & \langle X\rangle +0.35\left[\frac{\Delta t}{(1+z)9.8\ {\rm hr}}\right]
                            \left(\frac{Z_{\rm CNO}}{0.02}\right) \nonumber \\
    & = & z\frac{142}{\alpha}\frac{\xi_b}{\xi_p} - 0.145 
                           +\left[\frac{\Delta t}{(1+z)28\ {\rm hr}}\right]
                            \left(\frac{Z_{\rm CNO}}{0.02}\right)
\end{eqnarray}
Clearly this estimate of $X_0$ will incorporate systematic errors arising from the assumed values of $z$, the inclination, and $Z_{\rm CNO}$. We explore the typical range of these errors in 
% the next section.
\S\ref{sec:simulations}

>>>>

\usepackage{amsmath}	% to enable cases environment

@@

% Custom definitions

\newcommand{\eps}{{\rm erg\,s^{-1}}}
\newcommand{\epcs}{{\rm erg\,cm^{-2}\,s^{-1}}}
\newcommand{\epc}{{\rm erg\,cm^{-2}}}

@@

% \title{Thermonuclear burst model-observation comparisons with {\sc concord}}
\title{Robust inference of neutron-star parameters from thermonuclear burst 
% lightcurves}
observations}

@@

The fundamental observable that we measure is the time-history of the burst flux, the lightcurve $(t_i,F_{i})$. Here the $F_{i}$ may be in instrumental units of count~s$^{-1}$, or may be estimates of the bolometric flux at time $t_i$, derived from model fits of the spectrum at time $t_i$.
%
Provided the $F_i$ are in units of flux, or the flux can be estimated from the values,  we may measure the peak flux, $F_{\rm pk}$, and integrate to estimate the burst fluence, $E_b$.

The peak flux $F_{\rm pk}$ provides constraints on the source distance, since the maximum burst luminosity is limited to (roughly) the Eddington luminosity \cite[e.g.][]{lew93}
% copied from bcatalog
\begin{eqnarray}
  L_{\rm Edd,\infty} & = & \frac{8\pi G m_p M_{\rm NS} c
  [1+(\alpha_{\rm T}T_{\rm e})^{0.86}]} {\sigma_T(1+X)[1+z(R)]} % \xi_b
       \nonumber \\
  & = & 2.7\times10^{38} \left(\frac{M_{\rm NS}}{1.4M_\odot}\right)
 \frac{1+(\alpha_{\rm T}T_{\rm e})^{0.86}}{(1+X)}
\nonumber \\ & & \times\  % line break
    \left[\frac{1+z(R)}{1.31}\right]^{-1}\ % \xi_b\
              \eps
  \label{ledd}
\end{eqnarray}
Here
$T_{\rm e}$ is the
effective temperature of the atmosphere, $\alpha_{\rm T}$ parametrizes the temperature dependence of the electron scattering
opacity \cite[$\simeq 2.2\times10^{-9}$~K$^{-1}$;][]{lew93},
$m_p$ is the proton mass, $\sigma_T$ the Thompson 
cross-section, and $X$ the
mass fraction of hydrogen in the atmosphere.
% ($\approx0.7$ for cosmic abundances).
The final factor in
square brackets represents the gravitational redshift 
at the photosphere $1+z(R)=(1-2GM_{\rm NS}/R
c^2)^{-1/2}$, which may be lower than the value at the NS surface, while the photosphere is expanded during the radius expansion episode (i.e. $R\ge R_{\rm NS}$).
%
Mention also the empirical value here...

A significant fraction of observed bursts exhibit characteristic variations in their spectral hardness (or the blackbody temperature and normalisation) around their maximum that indicate the presence of photospheric radius-expansion (PRE). By equating the maximum flux of these events with the Eddington luminosity, the distance to the source can be estimated as
\begin{eqnarray}
 d & = & \left(\frac{L_{\rm Edd,\infty}}{4\pi \xi_b F_{\rm pk, RE}}\right)^{1/2} \nonumber \\
   & = & 8.6
	\left( \frac{\xi_b F_{\rm pk, RE}}{3\times10^{-8}\ \epcs} \right)^{-1/2}
       	\left(\frac{M_{\rm NS}}{1.4M_\odot}\right)^{1/2}
\nonumber \\ & & \times\  % line break
	\left[\frac{1+z(R)}{1.31}\right]^{-1/2}
	(1+X)^{-1/2}\ {\rm kpc}
 \label{disteq}
\end{eqnarray}
% 
Here the $\xi_b$ factor accounts for the possible anisotropy of the burst emission, and is defined in the same sense as \citep{fuji88}, such that $L_b=4\pi d^2\xi_b F_{\rm pk}$. Since the distribution of the burst flux over the neutron star is likely different from that of the persistent emission, we define a different anisotropy factor $\xi_p$ relating the observed persistent flux $F_p$ to the total luminosity. Modelling \cite[]{he16}...

As is obvious from equation \ref{disteq}, several factors will contribute to statistical (and possibly systematic) uncertainties in distance estimates derived in this manner. First, the neutron star mass $M_{\rm NS}$ is unknown...
%
Second, the radius $R$ at which the redshift should be calculated may not be clear from the observations. In principle one could take the inferred blackbody nromalisation at the time of maximum flux, but for most PRE bursts the maximum flux is achieved close to the ``touchdown'' point, where the photosphere has returned (more or less) to the neutron star surface \cite[e.g.][]{gal06a}.
%
Third, the possible range/likely value for $X$...
%
Fourth, even if the inclination $i$ of the system is known, converting to the anisotropy factor requires modelling of the effect of the disk on the radiation field arising from the burst \cite[e.g.][]{he16}.

These issues may be avoided by adopting instead the empirical Eddington luminosity measured by \cite[]{kuul03a}...

The fluence (and the shape of the burst profile) is determined by the amount of accumulated fuel, and its composition. The nuclear energy generation rate $Q_{\rm nuc}$ depends on the composition; studies with the 1-D numerical code {\sc kepler} suggest that 
\begin{eqnarray}
Q_{\rm nuc} &= & 1.31+6.95\bar{X}-1.92\bar{X}^2 \nonumber \\
& \approx & 1.35 + 6.05 \bar{X}
\end{eqnarray}
\cite[]{goodwin19a} where $\bar{X}$ is the average H-fraction of the fuel layer at ignition.

Perhaps put in here some estimates of the fuel composition as a function of $X_0$ and recurrence time; could perhaps develop a ``prior'' for $\bar{X}$

Provided an estimate of $\bar{X}$ is available, we can estimate the burst column that has ignited as
\begin{eqnarray}
y & = & \frac{L_b d^2 (1+z)}{R_{\rm NS}^2 Q_{\rm nuc}}\ 
                                                            \nonumber \\
  & = & 3.0\times10^8 \left(\frac{\xi_b E_{\rm b}}{10^{-6}\ \epc}\right)
                      \left(\frac{d}{10\ {\rm kpc}}\right)^2\
             \left(\frac{Q_{\rm nuc}}{4.4\ {\rm MeV/nucleon}}\right)^{-1}
\nonumber \\ & & \times\  % line break
                      \left(\frac{1+z}{1.31}\right)
                      \left(\frac{R_{\rm NS}}{10\ {\rm km}}\right)^{-2}
                      {\rm g\,cm^{-2}}
\label{column}
\end{eqnarray}
{\it need to update prefactor, denominator in $Q_{\rm nuc}$ term here }

From data taken prior to the burst ignition we can measure the persistent flux $F_{\rm per}$, a measure of the accretion rate; and provided more than one burst is observed, we may also measure (or estimate) the recurrence time, $\Delta t$.

These parameters may be combined to estimate the $\alpha$-value, the ratio of the burst to persistent flux: 
\begin{equation}
\alpha = \frac{\Delta t F_{\rm per} c_{\rm bol}}{E_B}
\end{equation}
where $c_{\rm bol}$ is the bolometric correction giving the inverse fraction of total persistent flux emitted in the instrumental band.

@@

% The corrections described below
However, since the model is to be transformed by a different value of $(1+z)_{\rm fit}$, we must take care to consistently transform the model parameters in accordance with this choice.
%
We can achieve this objective as follows. The surface gravity $g$ adopted for the model, and the trial redshift $(1+z)_{\rm fit}$ imply a different neutron star mass and radius than adopted for the model: 
%
% We note that the combination of surface gravity $g$ (assumed for the model) and adopted $1+z$  uniquely specifies the neutron star mass $M_{\rm NS}$ and radius $R_{\rm NS}$:
\begin{eqnarray}
% as used in function calc_mr in burstclass.py
R_{\rm NS,fit} &=& c^2\frac{(1+z)_{\rm fit}^2-1}{2g(1+z)_{\rm fit}}\\
M_{\rm NS,fit} &=& \frac{gR_{\rm NS,fit}^2}{G(1+z)_{\rm fit}}
\end{eqnarray}
Thus, by identifying the optimal value of $(1+z)_{\rm fit}$ for comparison to a particular observation, we can constrain the mass and radius, at a fixed $g$. [{\bf so which one should I use? and should I make any other changes to the recipe?}]

@@

Given an estimate of $Q_{\rm nuc}$, the average hydrogen fraction $\langle X\rangle$ in the fuel layer at ignition has been estimated to date using the relation $Q_{\rm nuc}=1.6 + 4\langle X\rangle$~MeV/nucleon \cite[e.g.][and references therein]{gal03d}. This expression includes $\approx35$\% losses
attributed to neutrino emission \cite[]{fuji87}. However, the 35\% value applies only to $\beta$-decays in the much more extensive {\it rp}-process burning chain, and recent {\sc kepler} simulations indicate instead an approximation of 
% $Q{\rm nuc}=0.96+6.6\langle X\rangle$ 
$Q{\rm nuc}=1.35+6.05\langle X\rangle$ 
% (Goodwin et al. 2018, in preparation). 
\cite[]{goodwin19a}.
[{\it comment on improved quadratic expression}]
%
Thus, by substituting in equation \ref{qnuc} we can estimate 
\begin{equation}
\langle X\rangle 
% In [42]: c.c**2/(6.0455*u.MeV/c.m_p).to("m**2/s**2")                             
% Out[41]: <Quantity 155.20173377>
%% = \frac{c^2z/\alpha}{6.6\ {\rm Mev/nucleon}}\frac{\xi_b}{\xi_p} - 0.145
% = z\frac{142}{\alpha}\frac{\xi_b}{\xi_p} - 0.145 \label{xbar}
 = z\frac{155}{\alpha}\frac{\xi_b}{\xi_p} - 0.223 \label{xbar}
\end{equation}
[{\it comment on the degree of error introduced by ignoring anisotropy etc.}] 
%
Clearly this expression is only applicable for $\alpha$-values up to some limit, which we calculate as
\begin{equation}
%In [43]: c.c**2/(1.*u.MeV/c.m_p).to("m**2/s**2")/1.3455        
% Out[43]: <Quantity 697.34082607>
% \alpha \leq 980\, z \frac{\xi_b}{\xi_p}
\alpha \leq 697\, z \frac{\xi_b}{\xi_p}
\end{equation}
Interestingly, this limit is 
% $\approx70$\% 
%  c.c**2/(1.*u.MeV/c.m_p).to("m**2/s**2")/1.6                 
% Out[44]: <Quantity 586.42005092>
% In [45]: 697./586.                                                               
% Out[45]: 1.189
$\approx20$\%
%
larger than the equivalent value derived for the old expression for $Q_{\rm nuc}$. This result suggests that larger observed values of $\alpha$ may be accommodated without resorting to explanations including incomplete burning of burst fuel \cite[e.g.][]{bcatalog}.

@@

The hydrogen fraction at ignition $\langle X\rangle$ and the burst recurrence time may then be used to estimate the hydrogen fraction in the fuel, $X_0$. The H-fraction at the base of the fuel layer is reduced steadily by $\beta$-limited hot CNO burning, and will completely exhaust the accreted hydrogen in a time \cite[]{lampe16}:
\begin{equation}
t_{\rm CNO} =9.8\left(\frac{X_0}{0.7}\right) \left(\frac{Z_{\rm CNO}}{0.02}\right)^{-1}
\end{equation}
measured in the neutron star frame.
% updated 2019 Aug 26 with exhaustion case
The hydrogen fraction at the base will thus be $X_0[1-\Delta t/(1+z)t_{\rm CNO}]$, 
% [{\it should this expression include time dilation?}] -- yes
%
provided that $\Delta t < (1+z)t_{\rm CNO}$. Once $\Delta t$ exceeds the time to burn all the hydrogen at the base, the abundance there will be zero, and a growing layer of pure He fuel will develop. The average H-fraction in the layer for these two cases will be 
%so that the average H-fraction in the layer will be 
\begin{equation}
\langle X \rangle = \begin{cases}
%  X_0\left[1-0.5\frac{\Delta t}{(1+z)t_{\rm CNO}}\right] & \Delta t \leq (1+z)t_{\rm CNO} \\
%  0.5X_0\frac{(1+z)t_{\rm CNO}}{\Delta t} & \Delta t > (1+z)t_{\rm CNO}
  X_0(1-0.5f_{\rm burn}) & \Delta t \leq (1+z)t_{\rm CNO} \\
  0.5X_0/f_{\rm burn} & \Delta t > (1+z)t_{\rm CNO}
  \end{cases}
\end{equation}
where $f_{\rm burn} = \frac{\Delta t}{(1+z)t_{\rm CNO}}$ is the ratio of the recurrence time to the time to burn all the H.
%
We can combine this expression with equation \ref{xbar} to give
\begin{equation}
X_0 = \begin{cases}
%        & \langle X\rangle +0.35\left[\frac{\Delta t}{(1+z)9.8\ {\rm hr}}\right]
%                            \left(\frac{Z_{\rm CNO}}{0.02}\right) \nonumber \\
         z\frac{142}{\alpha}\frac{\xi_b}{\xi_p} - 0.145 
                           +\left[\frac{\Delta t}{(1+z)28\,{\rm hr}}\right]
                       \left(\frac{Z_{\rm CNO}}{0.02}\right) & f_{\rm burn} \leq 1\\
         \sqrt{ \frac{\Delta t}{(1+z)7\,{\rm hr}}\frac{Z_{\rm CNO}}{0.02}
                        \left(z\frac{142}{\alpha}\frac{\xi_b}{\xi_p} - 0.145\right) }
                        & f_{\rm burn} > 1
        \end{cases}
\end{equation}
Practically, the issue with these expressions is that calculating $f_{\rm burn}$ requires knowledge of $X_0$, which is the unknown we are trying to constrain. One approach is to adopt a trial value of $X_0$, calculate $f_{\rm burn}$ and hence an updated estimate of $X_0$ via equation \ref{x0}, and iterate until no further change in the estimate arises.

Clearly this estimate of $X_0$ will incorporate systematic errors arising from the assumed values of $z$, the inclination, and $Z_{\rm CNO}$. We explore the typical range of these errors in 
% the next section.
\S\ref{sec:simulations}
